\bf Eye Fitting Straight Lines in the Modern Era

### 2021-09-09 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
rough rough* draft. I need to read through it for readability yet. Also why are intro's so hard??

## Fitted Values

In addition to the participant drawn points, $(x_k, y_{k,drawn})$, and the ordinary least squares (OLS) regression fitted values, $(x_k, \hat y_{k,OLS})$, a regression equation with a slope based on the first principal component (PCA) was used to calculate fitted values, $(x_k, \hat y_{k,PCA})$.
          For each set of simulated data and parameter choice, the PCA regression equation was determined by using the princomp function in the stats package in base R to obtain the rotation of the coordinate axes from the first principal component (direction which captures the most variance).
          The estimated slope, $\hat\beta_{1,PCA}$, is determined by the ratio of the axis rotation in y and axis rotation in x of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of of the simulated points, $(\bar x_i, \bar y_i)$.
            Fitted values, $\hat y_{k,PCA}$ are then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$.
          \cref{fig:ols-vs-pca-example} illustrates the difference between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation with a slope calculated by the first principal component which minimizes the smallest distance of points from the line.

+ Because of the randomness in the stimulus generation process, on any given trial the actual slope of the linear regression line could differ from its prescribed value.

## Feedback Data

## Linear Trend Constraint Performance

## Smoothing Spline Trend Performance

@@

@@

+ Use the method.
+ Write R package.

>>>>

## Fitted Regression Lines

We compare the participant drawn line to two regression lines determined by ordinary least squares regression and regression based on the principal axis (i.e. Deming Regression). \cref{fig:ols-vs-pca-example} illustrates the difference between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation based on the principal axis which minimizes the Euclidean distance of points (orthogonal) from the line.

Due to the randomness in the data generation process, the actual slope of the linear regression line fit through the simulated points could differ from the predetermined slope.
Therefore, we fit an ordinary least squares (OLS) regression to each scatter-plot to obtain fitted values in 0.25 increments across the domain, $\hat y_{k,OLS}$ for $k = 1, ..., 4 x_{max} +1$.

The regression equation based on the principal axis was determined by using the `princomp` function in the stats package in base R to obtain the rotation of the coordinate axes from the first principal component (direction which captures the most variance).
The estimated slope, $\hat\beta_{1,PCA}$, is determined by the ratio of the axis rotation in y and axis rotation in x of the first principal component with the y-intercept, $\hat\beta_{0,PCA}$ calculated by the point-slope equation of a line using the mean of of the simulated points, $(\bar x_i, \bar y_i)$.
Fitted values, $\hat y_{k,PCA}$, are then obtained every 0.25 increment across the domain from the PCA regression equation, $\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k$.

```{r ols-vs-pca-example, fig.height = 3, fig.width = 6, fig.cap="OLS vs PCA Regression Lines", message=FALSE, warning=FALSE, out.width="100%"}
library(ggplot2)
library(magrittr)
library(plyr)

set.seed(2)
corrCoef = 0.5 # sample from a multivariate normal, 10 datapoints
dat = MASS::mvrnorm(10,c(0,0),Sigma = matrix(c(1,corrCoef,2,corrCoef),2,2))
dat[,1] = dat[,1] - mean(dat[,1]) # it makes life easier for the princomp
dat[,2] = dat[,2] - mean(dat[,2])

dat = data.frame(x1 = dat[,1],x2 = dat[,2])

# Calculate the first principle component
# see http://stats.stackexchange.com/questions/13152/how-to-perform-orthogonal-regression-total-least-squares-via-pca
v = dat%>%prcomp%$%rotation
x1x2cor = bCor = v[2,1]/v[1,1]

x1tox2 = coef(lm(x1~x2,dat))
x2tox1 = coef(lm(x2~x1,dat))
slopeData = data.frame(slope = c(x1x2cor,x2tox1[2]),
                       type=c("PC", "OLS"))

# We want this to draw the neat orthogonal lines.
pointOnLine = function(inp){
  # y = a*x + c (c=0)
  # yOrth = -(1/a)*x + d
  # yOrth = b*x + d
  x0 = inp[1] 
  y0 = inp[2] 
  a = x1x2cor
  b = -(1/a)
  c = 0
  d = y0 - b*x0
  x = (d-c)/(a-b)
  y = -(1/a)*x+d
  return(c(x,y))
}

points = apply(dat,1,FUN=pointOnLine)

segmeData = rbind(data.frame(x=dat[,1],y=dat[,2],xend=points[1,],yend=points[2,],type = "PC"),
                  data.frame(x=dat[,1],y=dat[,2],yend=dat[,1]*x2tox1[2],xend=dat[,1],type="OLS"))

dat %>%
ggplot(aes(x1,x2))+
  geom_point()+
  geom_abline(data=slopeData,aes(slope = slope,intercept=0,color=type, linetype=type), size = 1.2)+
  geom_segment(data=segmeData,aes(x=x,y=y,xend=xend,yend=yend,color=type, linetype=type))+
  facet_grid(.~type)+
  coord_equal()+
  scale_x_continuous("x") +
  scale_y_continuous("y") +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "none",
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_blank(),
        # legend.text  = element_text(size = 10),
        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
        # strip.background = element_rect(size = 0.8),
        legend.key.size = unit(1, "line")
        ) +
  scale_color_manual(values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
  scale_linetype_manual(values = c("solid", "dashed"), labels = c("OLS", "PCA"))
```

## Data Availability

```{r eyefitting-plots}
eyefitting_model_data <- read.csv("data/youdrawit-eyefitting-model-data.csv") %>%
  dplyr::rename(`Parameter Choice` = parm_id)
```

For each participant, the final data set used for analysis contains $x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ for parameter choice $i = 1,2,3,4$, j = $1,...N_{participant}$, and $x_{ijk}$ value $k = 1, ...,4 x_{max} + 1$. 
Using both a linear mixed model and a generalized additive mixed model, comparisons of vertical residuals in relation to the OLS fitted values ($e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}$) and PCA fitted values ($e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}$) were made across the domain.
\cref{fig:eyefitting-example-plot} displays an example of all three fitted trend lines for parameter choice F.

```{r eyefitting-example-plot, fig.height = 4, fig.width = 4, fig.cap = "Eye Fitting Straight Lines in the Modern Era Example", out.width="65%"}
trial.feedback <- read.csv("data/youdrawit-eyefitting-example-feedback.csv") %>%
    mutate(`Parameter Choice` = "F")
trial.sim <- read.csv("data/youdrawit-eyefitting-example-simulated.csv") %>%
    mutate(`Parameter Choice` = "F")

trial.feedback %>%
  ggplot(aes(x = x)) +
  geom_line(aes(y = y, color = "OLS", linetype = "OLS")) +
  geom_line(aes(y = ypca, color = "PCA", linetype = "PCA")) +
  geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
  geom_point(data = trial.sim, aes(y = y)) +
  facet_wrap(~`Parameter Choice`) +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "bottom",
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
        # strip.background = element_rect(size = 0.8),
        legend.key.size = unit(1, "line")
        ) +
  scale_x_continuous(limits = c(0,20)) +
  scale_y_continuous(limits = c(-5, 17), breaks = seq(-5, 15, 5)) +
  scale_color_manual("", values = c("black", "steelblue", "orange")) +
  scale_linetype_manual("", values = c("dashed", "solid", "solid"))
```

## Linear Trend Constraint

Using the lmer function in the lme4 package [@lme4], a linear mixed model (LMM) is fit separately to the OLS and PCA residuals, constraining the fit to a linear trend. 
Parameter choice, $x$, and the interaction between $x$ and parameter choice were treated as fixed effects with a random participant effect accounting for variation due to participant.
The LMM equation for each fit (OLS and PCA) residuals is given by:
\begin{equation}
y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
\end{equation}
\noindent where

+ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value
+ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
+ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
+ $\gamma_0$ is the overall intercept
+ $\alpha_i$ is the effect of the $i^{th}$ parameter choice (F, S, V, N) on the intercept
+ $\gamma_1$ is the overall slope for $x$
+ $\gamma_{2i}$ is the effect of the parameter choice on the slope
+ $x_{ijk}$ is the x-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
+ $p_{j} \sim N(0, \sigma^2_{participant})$ is the random error due to the $j^{th}$ participant's characteristics
+ $\epsilon_{ijk} \sim N(0, \sigma^2)$ is the residual error.

```{r eyefitting-lmer-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Eye Fitting Straight Lines in the Modern Era LMM results"}
eyefitting.preds.lmer <- read.csv("data/youdrawit-eyefitting-lmerpred-data.csv")
# Plot Predictions
eyefitting.lmer.plot <- eyefitting.preds.lmer %>%
  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
  dplyr::rename(`Parameter Choice` = parm_id) %>%
  ggplot(aes(x = x)) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.7) +
  geom_line(aes(y = emmean.ols, color = "OLS")) +
  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PCA"), color = NA, alpha = 0.7) +
  geom_line(aes(y = emmean.pca, color = "PCA")) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~`Parameter Choice`, ncol = 4) +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "right",
        plot.title   = element_text(size = 12, hjust = 0),
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
        # strip.background = element_rect(size = 0.5),
        legend.key.size = unit(1, "line")
        ) +
  scale_y_continuous("Residual") +
  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
  scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))

eyefitting.lmer.plot
```

## Smoothing Spline Trend

Eliminating the linear trend constraint, the bam function in the mgcv package [@mgcv1; @mgcv2; @mgcv3; @mgcv4; @mgcv5] is used to fit a generalized additive mixed model (GAMM) separately to the OLS and PCA residuals to allow for estimation of smoothing splines.
Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice. A random participant effect accounting for variation due to participant and a random spline for each participant accounted for variation in spline for each participant.
The GAMM equation for each fit (OLS and PCA) residuals is given by:
\begin{equation}
y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
\end{equation}
\noindent where

+ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value
+ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
+ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of x-value corresponding to either the OLS or PCA fit
+ $\alpha_i$ is the intercept for the parameter choice $i$
+ $s_{i}$ is the smoothing spline for the $i^{th}$ parameter choice
+ $x_{ijk}$ is the x-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
+ $p_{j} \sim N(0, \sigma^2_{participant})$ is the error due to participant variation
+ $s_{j}$ is the random smoothing spline for each participant.

\cref{fig:eyefitting-lmer-residualplots} and \cref{fig:eyefitting-gamm-residualplots} show the estimated trends of residuals (vertical deviation of participant drawn points from both the OLS and PCA fitted points) as modeled by a LMM and GAMM respectively. 
Examining the plots, the estimated trends of PCA residuals (orange) appear to align closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue). 
In particular, this trend is more prominent in parameter choices with large variances (F and N).
These results are consistent to those found in @mosteller1981eye indicating participants fit a trend line closer to the estimated regression line with the slope of the first principal component than the estimated OLS regression line.

```{r eyefitting-gamm-residualplots, fig.height = 3, fig.width = 9, out.width = "100%", fig.cap = "Eye Fitting Straight Lines in the Modern Era GAMM results"}
eyefitting.grid.gamm <- read.csv("data/youdrawit-eyefitting-gammpred-data.csv")
eyefitting.gamm.plot <- eyefitting.grid.gamm %>%
  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
  dplyr::rename(`Parameter Choice` = parm_id) %>%
  ggplot(aes(x = x)) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualols, group = plotID, color = "OLS"), alpha = 0.1) +
  geom_line(data = eyefitting_model_data, aes(x = x, y = residualpca, group = plotID, color = "PCA"), alpha = 0.1) +
  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.5) +
  geom_line(aes(y = ols.pred, color = "OLS")) +
  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PCA"), color = NA, alpha = 0.5) +
  geom_line(aes(y = pca.pred, color = "PCA")) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~`Parameter Choice`, ncol = 4) +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "right",
        plot.title   = element_text(size = 12, hjust = 0),
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
        # strip.background = element_rect(size = 0.5),
        legend.key.size = unit(1, "line")
        ) +
  scale_y_continuous("Residual") +
  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
  scale_fill_manual("GAMM fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) 
eyefitting.gamm.plot
```

@@

The intent of research was to establish you draw it as a tool for testing graphics then use this tool to determine the cognitive implications of displaying data on the log scale.
Eye Fitting Straight Lines in the Modern Era replicated the results found in @mosteller1981eye. 
When shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least squares regression line.
This trend was most prominent when shown data simulated with larger variances. 
The reproducibility of these results serve as evidence of the reliability of the you draw it method.

@@

This study provided a basis for the use of 'You Draw It' as a tool for testing graphics.
Further investigation is necessary to implement the method in non-linear settings and with real data.
This tool could also be used to evaluate human ability to extrapolate data from trends. 
In the future, an R package designed for easy implementation of 'You Draw It' task plots will help make this tool accessible to other researchers.

### 2021-09-09 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
merged. thanks for catching spelling error! Whoops :)
rough rough* draft. I need to read through it for readability yet. Also why are intro's so hard??

{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-simplot-1}

@@

\hypertarget{fitted-values}{%
\subsection{Fitted Values}\label{fitted-values}}

In addition to the participant drawn points, \((x_k, y_{k,drawn})\), and
the ordinary least squares (OLS) regression fitted values,
\((x_k, \hat y_{k,OLS})\), a regression equation with a slope based on
the first principal component (PCA) was used to calculate fitted values,
\((x_k, \hat y_{k,PCA})\). For each set of simulated data and parameter
choice, the PCA regression equation was determined by using the princomp
function in the stats package in base R to obtain the rotation of the
coordinate axes from the first principal component (direction which
captures the most variance). The estimated slope, \(\hat\beta_{1,PCA}\),
is determined by the ratio of the axis rotation in y and axis rotation
in x of the first principal component with the y-intercept,
\(\hat\beta_{0,PCA}\) calculated by the point-slope equation of a line
using the mean of of the simulated points, \((\bar x_i, \bar y_i)\).
Fitted values, \(\hat y_{k,PCA}\) are then obtained every 0.25 increment
across the domain from the PCA regression equation,
\(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).
\cref{fig:ols-vs-pca-example} illustrates the difference between an OLS
regression equation which minimizes the vertical distance of points from
the line and a regression equation with a slope calculated by the first
principal component which minimizes the smallest distance of points from
the line.

\begin{itemize}
\tightlist
\item
  Because of the randomness in the stimulus generation process, on any
  given trial the actual slope of the linear regression line could
  differ from its prescribed value.
\end{itemize}

\hypertarget{feedback-data}{%
\subsection{Feedback Data}\label{feedback-data}}

\hypertarget{linear-trend-constraint-performance}{%
\subsection{Linear Trend Constraint
Performance}\label{linear-trend-constraint-performance}}

\hypertarget{smoothing-spline-trend-performance}{%
\subsection{Smoothing Spline Trend
Performance}\label{smoothing-spline-trend-performance}}

@@

@@

\begin{itemize}
\tightlist
\item
  Use the method.
\item
  Write R package.
\end{itemize}

>>>>

{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-simplot-1}

@@

\hypertarget{fitted-regression-lines}{%
\subsection{Fitted Regression Lines}\label{fitted-regression-lines}}

We compare the participant drawn line to two regression lines determined
by ordinary least squares regression and regression based on the
principal axis (i.e.~Deming Regression). \cref{fig:ols-vs-pca-example}
illustrates the difference between an OLS regression equation which
minimizes the vertical distance of points from the line and a regression
equation based on the principal axis which minimizes the Euclidean
distance of points (orthogonal) from the line.

Due to the randomness in the data generation process, the actual slope
of the linear regression line fit through the simulated points could
differ from the predetermined slope. Therefore, we fit an ordinary least
squares (OLS) regression to each scatter-plot to obtain fitted values in
0.25 increments across the domain, \(\hat y_{k,OLS}\) for
\(k = 1, ..., 4 x_{max} +1\).

The regression equation based on the principal axis was determined by
using the \texttt{princomp} function in the stats package in base R to
obtain the rotation of the coordinate axes from the first principal
component (direction which captures the most variance). The estimated
slope, \(\hat\beta_{1,PCA}\), is determined by the ratio of the axis
rotation in y and axis rotation in x of the first principal component
with the y-intercept, \(\hat\beta_{0,PCA}\) calculated by the
point-slope equation of a line using the mean of of the simulated
points, \((\bar x_i, \bar y_i)\). Fitted values, \(\hat y_{k,PCA}\), are
then obtained every 0.25 increment across the domain from the PCA
regression equation,
\(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).

{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/ols-vs-pca-example-1}

\caption{OLS vs PCA Regression Lines}\label{fig:ols-vs-pca-example}
\end{figure}

\hypertarget{data-availability}{%
\subsection{Data Availability}\label{data-availability}}

For each participant, the final data set used for analysis contains
\(x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}\), and \(\hat y_{ijk,PCA}\)
for parameter choice \(i = 1,2,3,4\), j = \(1,...N_{participant}\), and
\(x_{ijk}\) value \(k = 1, ...,4 x_{max} + 1\). Using both a linear
mixed model and a generalized additive mixed model, comparisons of
vertical residuals in relation to the OLS fitted values
(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PCA fitted
values (\(e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}\)) were made
across the domain. \cref{fig:eyefitting-example-plot} displays an
example of all three fitted trend lines for parameter choice F.

{\centering \includegraphics[width=0.65\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-example-plot-1}

\caption{Eye Fitting Straight Lines in the Modern Era Example}\label{fig:eyefitting-example-plot}
\end{figure}

\hypertarget{linear-trend-constraint}{%
\subsection{Linear Trend Constraint}\label{linear-trend-constraint}}

Using the lmer function in the lme4 package \citep{lme4}, a linear mixed
model (LMM) is fit separately to the OLS and PCA residuals, constraining
the fit to a linear trend. Parameter choice, \(x\), and the interaction
between \(x\) and parameter choice were treated as fixed effects with a
random participant effect accounting for variation due to participant.
The LMM equation for each fit (OLS and PCA) residuals is given by:
\begin{equation}
y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
\end{equation} \noindent where

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
  choice, \(j^{th}\) participant, and \(k^{th}\) increment of x-value
\item
  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  x-value corresponding to either the OLS or PCA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
  \(k^{th}\) increment of x-value corresponding to either the OLS or PCA
  fit
\item
  \(\gamma_0\) is the overall intercept
\item
  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (F, S,
  V, N) on the intercept
\item
  \(\gamma_1\) is the overall slope for \(x\)
\item
  \(\gamma_{2i}\) is the effect of the parameter choice on the slope
\item
  \(x_{ijk}\) is the x-value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the random error due to
  the \(j^{th}\) participant's characteristics
\item
  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
\end{itemize}

{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-lmer-residualplots-1}

\caption{Eye Fitting Straight Lines in the Modern Era LMM results}\label{fig:eyefitting-lmer-residualplots}
\end{figure}

\hypertarget{smoothing-spline-trend}{%
\subsection{Smoothing Spline Trend}\label{smoothing-spline-trend}}

Eliminating the linear trend constraint, the bam function in the mgcv
package \citep{mgcv1, mgcv2, mgcv3, mgcv4, mgcv5} is used to fit a
generalized additive mixed model (GAMM) separately to the OLS and PCA
residuals to allow for estimation of smoothing splines. Parameter choice
was treated as a fixed effect with no estimated intercept and a separate
smoothing spline for \(x\) was estimated for each parameter choice. A
random participant effect accounting for variation due to participant
and a random spline for each participant accounted for variation in
spline for each participant. The GAMM equation for each fit (OLS and
PCA) residuals is given by: \begin{equation}
y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
\end{equation} \noindent where

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
  choice, \(j^{th}\) participant, and \(k^{th}\) increment of x-value
\item
  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  x-value corresponding to either the OLS or PCA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
  \(k^{th}\) increment of x-value corresponding to either the OLS or PCA
  fit
\item
  \(\alpha_i\) is the intercept for the parameter choice \(i\)
\item
  \(s_{i}\) is the smoothing spline for the \(i^{th}\) parameter choice
\item
  \(x_{ijk}\) is the x-value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the error due to
  participant variation
\item
  \(s_{j}\) is the random smoothing spline for each participant.
\end{itemize}

\cref{fig:eyefitting-lmer-residualplots} and
\cref{fig:eyefitting-gamm-residualplots} show the estimated trends of
residuals (vertical deviation of participant drawn points from both the
OLS and PCA fitted points) as modeled by a LMM and GAMM respectively.
Examining the plots, the estimated trends of PCA residuals (orange)
appear to align closer to the \(y=0\) horizontal (dashed) line than the
OLS residuals (blue). In particular, this trend is more prominent in
parameter choices with large variances (F and N). These results are
consistent to those found in \citet{mosteller1981eye} indicating
participants fit a trend line closer to the estimated regression line
with the slope of the first principal component than the estimated OLS
regression line.

{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-gamm-residualplots-1}

\caption{Eye Fitting Straight Lines in the Modern Era GAMM results}\label{fig:eyefitting-gamm-residualplots}
\end{figure}

@@

The intent of research was to establish you draw it as a tool for
testing graphics then use this tool to determine the cognitive
implications of displaying data on the log scale. Eye Fitting Straight
Lines in the Modern Era replicated the results found in
\citet{mosteller1981eye}. When shown points following a linear trend,
participants tended to fit the slope of the first principal component
over the slope of the least squares regression line. This trend was most
prominent when shown data simulated with larger variances. The
reproducibility of these results serve as evidence of the reliability of
the you draw it method.

@@

This study provided a basis for the use of `You Draw It' as a tool for
testing graphics. Further investigation is necessary to implement the
method in non-linear settings and with real data. This tool could also
be used to evaluate human ability to extrapolate data from trends. In
the future, an R package designed for easy implementation of `You Draw
It' task plots will help make this tool accessible to other researchers.
